---
title: "ALFAZ MIDTERM FILE"
author: "Alfazkhan"
date: "2024-11-04"
output: html_document
---

```{r}
library(readr)
library(fpp2)
library(ggplot2)
library(forecast)

```{r}

# Import the dataset
sales_data <- read.csv("C:/Users/Alfaz Khan/Downloads/sales.csv", header = FALSE)
colnames(sales_data) <- c("Date", "Sales")

# Convert Sales column to a time series object
ts_sales <- ts(sales_data$Sales, start = c(2020, 1), frequency = 12) 
plot(ts_sales)

# Observations of the time series plot:
# Trend: The series shows an overall increasing/decreasing (choose one) trend over time.
# Seasonality: Consistent peaks/troughs observed around specific months, indicating seasonality.
# Volatility: Significant month-to-month fluctuations suggest variability in sales demand.
# Stationarity: The series appears non-stationary with changing mean/variance, suggesting the need for transformations.
# Accuracy Measure: Due to the observed fluctuations and non-stationary behavior in sales, MAPE is used to assess forecast accuracy.
# MAPE, by expressing forecast errors as percentages, provides insights into relative error and helps assess the model's effectiveness in dynamic data.

summary(ts_sales)

# Box plot of the time series data
boxplot(ts_sales, main = "Boxplot of Sales Data", ylab = "Sales")

# Explanation of Summary and Box Plot for Sales Data
# The summary statistics reveal the overall range, average, and central tendency of sales values.
# Min and Max indicate the lowest and highest sales periods, helping identify range and variability.
# The 1st and 3rd quartiles show the middle spread, suggesting the typical fluctuation in sales.
# The box plot visually shows distribution and highlights any potential outliers, indicating unusual sales patterns.

# Decompose the series
sales_decomposition <- stl(ts_sales, s.window = "periodic")
plot(sales_decomposition)

# Explanation of Decomposition:
# The decomposition separates the series into trend, seasonal, and remainder components.
# Trend: This component shows the underlying direction of the data.It is upwards and slightly dips after reaching it's peak in 2023.
# Seasonal: The seasonal component reveals recurring patterns within each period (monthly seasonality).
# Remainder: The remainder (or residual) component shows random fluctuations not explained by trend or seasonality.

# STL Decomposition Type:
# Based on the decomposition, the seasonal component shows consistent fluctuations over time.
# The seasonal effect does not increase with the trend, indicating an additive model is appropriate.
# In an additive model, the seasonal fluctuations are constant in magnitude, fitting well with the stable seasonal pattern seen here.
# Thus, additive decomposition is used as it accurately represents the seasonal pattern in this data.

# Seasonal indices
seasonal_indices <- sales_decomposition$time.series[, "seasonal"]
print(seasonal_indices)

# The seasonal component has monthly indices that describe typical deviations from the trend each month.
# High/Low Seasonal Values: The month with the highest seasonal value indicates when sales peak, while the lowest shows a dip in sales.

# Identify high and low seasonal months
high_value <- max(seasonal_indices)
low_value <- min(seasonal_indices)
high_month <- which(seasonal_indices == high_value)[1]
low_month <- which(seasonal_indices == low_value)[1]

cat("High sales month:", high_month, "with value:", high_value, "\n")
cat("Low sales month:", low_month, "with value:", low_value, "\n")

# Explanation for high and low seasonal months
# High months often align with peak demand periods, such as holiday seasons or end-of-year sales.
# Low months might reflect off-peak demand periods, perhaps due to seasonally low activity or market factors.

# Adjusting for seasonality
seasonally_adjusted_sales <- ts_sales - seasonal_indices
plot(ts_sales, col = "blue", main = "Actual vs. Seasonally Adjusted Sales")
lines(seasonally_adjusted_sales, col = "red", lty = 2)

# Explanation of Seasonally Adjusted Plot:
# The overlay shows actual sales and seasonally adjusted sales. Differences highlight the effect of seasonality.
# Large fluctuations between the actual and seasonally adjusted series indicate a strong seasonal influence.
# Removing seasonality can help focus on the trend and irregular fluctuations for forecasting.



# Naive forecast method on time series data
naive_forecast <- naive(ts_sales, h=12)
plot(naive_forecast)

# Explanation of Naive Method:
# The Naive method assumes that future values will be the same as the last observed value.
# It serves as a simple benchmark, offering a straightforward way to make forecasts with minimal calculation.
# Particularly useful for data with random walks or non-trending time series.

# Residual Analysis:
plot(naive_forecast$residuals, main = "Residuals of Naive Forecast", ylab = "Residuals", xlab = "Time")
# Plot of Residuals:
# Visual inspection of residuals helps identify any systematic patterns left unexplained by the model.
# A random scatter around zero without clear patterns indicates that the Naive model has captured the structure in the data.

# Histogram of residuals
hist(naive_forecast$residuals, main = "Histogram of Residuals", xlab = "Residuals")
# Explanation of Histogram:
# The histogram checks if residuals are roughly normally distributed.
# A symmetric distribution around zero indicates balanced forecast errors.

# ACF plot of residuals
Acf(naive_forecast$residuals, main = "ACF of Residuals")
# The ACF (Autocorrelation Function) plot examines correlations among residuals at different lags.
# Lack of significant autocorrelations indicates residuals are uncorrelated, suggesting a good model fit.

# Calculate accuracy metrics for the Naive forecast
accuracy_metrics <- accuracy(naive_forecast)
print(accuracy_metrics)

naive_forecast_next_year <- forecast(naive_forecast)
plot(naive_forecast_next_year)
# Using the Naive method, we project the future values by maintaining the last observed value as the prediction for each future time point.
# Summary of the Naive Forecasting Technique

# The Naive method is a simple forecasting approach where each future value is set equal to the last observed value.
# This method serves as a basic benchmark, particularly useful for time series data without strong trends or seasonal patterns.

# How good is the accuracy?
# Based on the accuracy metrics:
# RMSE (Root Mean Squared Error) reflects the average size of the forecast error, showing moderate error levels.
# MAE (Mean Absolute Error) provides the average absolute forecast error, indicating deviations from actual values.
# MAPE (Mean Absolute Percentage Error) expresses the error as a percentage, useful for understanding relative forecast accuracy.
# ACF1 (First-order autocorrelation of residuals) shows low to moderate autocorrelation in residuals, suggesting that the model does not systematically under- or over-forecast.
# These metrics suggest that while the Naive method provides a straightforward forecast, it might not be the most accurate for data with patterns.

# What does it predict the time series value will be in one year?
# The Naive forecast projects that the value will remain constant at the last observed sales figure for the next 12 months.
# This results in a stable, flat forecast line for the next year, assuming no changes or trends in the data.

# Other Observations:
# The Naive forecast works as a benchmark model, offering a simple prediction that can be used to compare against more complex models.
# While it provides a stable prediction, the Naive approach may not capture underlying trends or seasonal patterns, leading to higher forecast errors if such patterns are present in the data.


# Simple Moving Average
sales_maf <- ma(ts_sales, order=12)
plot(sales_maf, main="Moving Average Forecasting")

# Simple Moving Average of order 3
ma3 <- ma(ts_sales, order = 3)
lines(ma3, col = "red", lty = 1)  # Adding MA(3) in red

# Simple Moving Average of order 6
ma6 <- ma(ts_sales, order = 6)
lines(ma6, col = "blue", lty = 2)  # Adding MA(6) in blue

# Simple Moving Average of order 9
ma9 <- ma(ts_sales, order = 9)
lines(ma9, col = "green", lty = 3)  # Adding MA(9) in green

# Adding a legend to distinguish between different moving averages
legend("topright", legend = c("Original Series", "MA(3)", "MA(6)", "MA(9)"),
       col = c("black", "red", "blue", "green"), lty = c(1, 1, 2, 3))


# Observations on Moving Averages:
# As the moving average order increases (3 to 6 to 9), the plot becomes smoother, focusing on the long-term trend.
# Higher-order moving averages reduce noise but introduce a lag, capturing general trends while ignoring short-term fluctuations.




# Simple Smoothing Forecast
sales_ses <- ses(ts_sales, h = 12)
plot(sales_ses)

# Display summary to get alpha, initial state, and sigma all at once
summary(sales_ses)

# Explanation of Parameters and Model Output
# Alpha: The smoothing parameter, derived from the data, determines how much weight recent values carry.
# Here, alpha is set to 0.4319, indicating a moderate level of responsiveness to recent values.
# Initial State: The initial level of the series, found in the summary output, starts the forecast from a base level.
# Sigma: Represents the residual standard deviation, indicating forecast uncertainty. Lower sigma suggests more reliable forecasts.


# Plot of Residuals
plot(sales_ses$residuals, main = "Residuals of Simple Smoothing Forecast", ylab = "Residuals", xlab = "Time")
# This plot helps visualize any noticeable patterns in residuals. 
# A random scatter around zero indicates the model has captured most patterns in the data, 
# while systematic patterns suggest the model may miss some trends.

# Histogram of Residuals
hist(sales_ses$residuals, main = "Histogram of Residuals", xlab = "Residuals")
# Shows the distribution of residuals. A symmetric distribution around zero suggests balanced forecast errors 
# with minimal bias in the model.

# Plot of Fitted Values vs Residuals
plot(fitted(sales_ses), sales_ses$residuals, main = "Fitted Values vs Residuals", xlab = "Fitted Values", ylab = "Residuals")
# Checks for correlation between fitted values and residuals. A random spread around zero suggests a good fit. 
# Patterns here could indicate systematic errors.

# Plot of Actual Values vs Residuals
plot(ts_sales, sales_ses$residuals, main = "Actual Values vs Residuals", xlab = "Actual Sales", ylab = "Residuals")
# Shows if residuals vary with actual values. A random distribution around zero suggests residual independence, supporting model adequacy.

# ACF Plot of Residuals
Acf(sales_ses$residuals, main = "ACF of Residuals for Simple Smoothing Forecast")
# Examines residual correlations at different lags. Lack of significant autocorrelation (spikes close to zero) 
# indicates uncorrelated residuals, which supports a well-fitted model.

# Display accuracy measures for the Simple Smoothing forecast
accuracy(sales_ses)
# The accuracy metrics (RMSE, MAE, MAPE) provide insights into the model's precision.
# Moderate RMSE and MAE suggest a fair fit, though improvement might be possible with models accounting for trends.
# A low ACF1 value shows minimal autocorrelation in residuals, supporting the model’s adequacy.

# Forecast for the Next 12 Months and Explanation
sales_ses_forecast <- forecast(sales_ses, h = 12)
print(sales_ses_forecast)
plot(sales_ses_forecast, main = "Simple Smoothing Forecast for Next Year")
# The forecast plot for the next 12 months includes shaded areas representing 80% and 95% confidence intervals,
# offering a range within which future values are expected, thus indicating forecast uncertainty.

# Summary of Simple Smoothing Technique
# Interpretation of Forecast: The forecast suggests sales will stabilize around the last observed level, with minor fluctuations.
# Predicted values imply stability in the coming year, assuming no new trends or seasonal changes.
# Simple Exponential Smoothing is suitable for relatively stable data without strong trends or seasonality.
# Effective for short-term forecasting but may not fully capture complex patterns like seasonality or trends.



# Holt-Winters Forecasting
sales_hw <- HoltWinters(ts_sales, seasonal = "additive")
sales_hw_forecast <- forecast(sales_hw, h = 12)
plot(sales_hw_forecast, main = "Holt-Winters Forecast for Next 12 Months")

summary(sales_hw)
# Explanation of Holt-Winters Model Parameters and Output:
# Alpha : Controls the smoothing of the level. Higher alpha makes the forecast more responsive to recent changes.
# Beta : Adjusts the trend's smoothness. Higher beta values make the trend more sensitive to recent data.
# Gamma : Determines the weight of seasonal patterns. Higher gamma adapts more quickly to seasonal changes.
# Initial States: Level, trend, and seasonal starting points derived from the data, which serve as the base for forecasting.
# Sigma: The standard deviation of residuals, indicating the uncertainty in the forecast. Lower sigma values suggest a more stable model.

# Plot Residuals
plot(sales_hw_forecast$residuals, main = "Residuals of Holt-Winters Forecast", ylab = "Residuals", xlab = "Time")
# Explanation: Random scatter around zero with no obvious patterns indicates a well-fitting model.

# Histogram of Residuals
hist(sales_hw_forecast$residuals, main = "Histogram of Residuals", xlab = "Residuals")
# Explanation: A symmetric histogram centered around zero suggests balanced forecast errors and minimal bias.

# Fitted Values vs. Residuals Plot
plot(fitted(sales_hw_forecast), sales_hw_forecast$residuals, main = "Fitted Values vs Residuals", xlab = "Fitted Values", ylab = "Residuals")
# Explanation: Ideally, this plot should display no systematic patterns, confirming an adequate model fit.

# Actual Values vs. Residuals Plot
plot(ts_sales, sales_hw_forecast$residuals, main = "Actual Values vs Residuals", xlab = "Actual Sales", ylab = "Residuals")
# Explanation: A random distribution around zero implies residual independence and supports the model’s reliability.

# ACF Plot of Residuals
Acf(sales_hw_forecast$residuals, main = "ACF of Residuals for Holt-Winters Forecast")
# Explanation: Lack of significant autocorrelations in residuals suggests that the model captures the structure of the data well.

# Print accuracy measures for Holt-Winters forecast
accuracy(sales_hw_forecast)
# Explanation: Provides RMSE, MAE, and MAPE as measures of forecast accuracy. These help evaluate how closely the forecast matches the actual values.

# Forecast Table and Explanation
print(sales_hw_forecast)
plot(sales_hw_forecast, main = "Holt-Winters Forecast", ylab = "Sales", xlab = "Time")

# Forecast Interpretation
# This plot visualizes the Holt-Winters forecast for the upcoming 12 months, accounting for trend and seasonality.
# Shaded areas represent forecast intervals (80% and 95%), indicating the expected range of future values.
# Forecast accuracy measures (RMSE, MAE, MAPE) provide insights into the model's predictive performance.
# Lower residual autocorrelation (ACF1) and low sigma values reinforce the model's effectiveness in capturing seasonal patterns and trends.



# Calculate MAPE for each model
naive_mape <- accuracy(naive_forecast)[, "MAPE"]
mean_mape <- accuracy(meanf(ts_sales, h=12))[, "MAPE"]
ses_mape <- accuracy(sales_ses)[, "MAPE"]
hw_mape <- accuracy(sales_hw_forecast)[, "MAPE"]

mape_table <- data.frame(
  Model = c("Naive", "Mean", "SES", "Holt-Winters"),
  MAPE = c(naive_mape, mean_mape, ses_mape, hw_mape)
)

mape_table <- mape_table[order(mape_table$MAPE), ]
print(mape_table)



# Forecasting Method Definitions and Usefulness

# 1. Naive Forecasting:
# Definition: Projects future values as identical to the most recent observed value.
# Usefulness: Effective for series with high randomness or when recent values best indicate future values.
#   Serves as a baseline for comparison against more sophisticated models.

# 2. Simple Moving Average (SMA):
# Definition: Calculates the average of a fixed number of past observations.
# Usefulness: Smooths out fluctuations to highlight longer-term trends, useful in stable series with minimal trend or seasonality.

# 3. Simple Exponential Smoothing (SES):
# Definition: Applies exponentially decreasing weights to past observations, focusing on recent data.
# Usefulness: Best for short-term forecasting when data lacks trend or seasonality, as it adapts to changes in level but not trend.

# 4. Holt-Winters Method:
# Definition: Extends exponential smoothing with trend and seasonal components (beta and gamma).
# Usefulness: Ideal for datasets with both trend and seasonal patterns, as it captures these changes over time.

# 5. STL Decomposition:
# Definition: Separates a time series into seasonal, trend, and residual components for seasonal adjustment before forecasting.
# Usefulness: Useful for handling complex seasonality patterns. STL decomposition clarifies trends and seasonal effects, allowing for accurate forecasts.

# Best and Worst Forecast Models for Each Accuracy Measure
# Best Forecasting Model: The Simple Exponential Smoothing (SES) model has the lowest MAPE (15.49), indicating it most accurately captures data patterns with minimal relative error in this dataset.
# Worst Forecasting Model: The Mean model has the highest MAPE (140.83), showing it is the least suitable as it exhibits substantial forecast errors relative to actual values.

# Interpretation:
# The SES model’s low MAPE reflects its ability to handle data with some variability but without strong seasonal or trend components.
# The high MAPE for the Mean model suggests it’s ineffective for data with trends or seasonality, as it doesn’t adapt to recent changes.

# Summary of Time Series Analysis:
# Over the analyzed period, the time series displayed significant variability with discernible patterns.
# Seasonal peaks and troughs were evident, likely due to cyclic demand periods. This highlights recurring trends in the data that align with specific times of the year.

# Forecasting Outlook for the Next Year and Beyond:
# Based on the forecast models, the series is expected to remain relatively stable over the next year, suggesting minor fluctuations but no sharp upward or downward trend.
# Over the next two years, if no new influencing factors arise, the data is projected to continue its seasonal pattern with minimal long-term trend.

# Ranking of Forecasting Methods:
# 1.Simple Exponential Smoothing (SES): This model achieved the best accuracy with the lowest MAPE, making it well-suited to capturing the series’ characteristics without strong seasonality.
# 2.Holt-Winters (HW): Provided a slightly higher MAPE than SES, but it effectively captured seasonality, proving useful for data with seasonal fluctuations.
# 3.Naive Method: As a baseline model, it delivered moderate accuracy, providing a simplistic approach that performed adequately for short-term stability.
# 4.Mean Method: This model had the highest MAPE and was the least effective, underscoring its limited suitability for data with trends and seasonality.

# Concluding Observations:
# The SES model stands out as the best fit for this time series, given its ability to adjust to short-term fluctuations. However, HW was beneficial for highlighting seasonality.
# Ultimately, the SES model offers the most accurate and reliable short-term forecast, while HW provides a comprehensive view of seasonal patterns.
